Towards Proof-Reconstruction of Problems in Classical Propositional Logic in Agda

### 2017-09-05 paper/sections/introduction.tex
[ paper ] new commands for programs and langs.

% -*- root: main.tex -*-
\documentclass[main.tex]{subfiles}
\begin{document}

@@

Proof reconstruction is a hard labor since it depends on the integration
of two complex systems. On the one hand, we have the automatic theorem provers
(henceforth ATP) and their specification logic. These tools are usually
classified in at least one of the following categories. A SAT solver
(e.g., \verb!zChaff!~\cite{Moskewicz2001} and \verb!MiniSat!~\cite{Een2004})
to prove unsatisfiability of CNF formulas, a QBF solver
(e.g., \verb!GhostQ!~\cite{Klieber2014} and \verb!DepQBF!~\cite{Lonsing2017})
to prove satisfiability and invalidity of quantified Boolean formulas, a SMT
solver (e.g \verb!CVC4!~\cite{Barrett2011}, \verb!veriT!~\cite{bouton2009},
and \verb!Z3!~\cite{DeMoura2008}) to prove unsatisfiability of formulas
from first-order logic with theories, and a prover for validity of
formulas from first-order logic with equality
(e.g., \verb!E!~\cite{Schulz:AICOM-2002}, \verb!leanCoP!~\cite{Otten2008},
\verb!Metis!~\cite{hurd2003first}, \verb!SPASS! ~\cite{Weidenbach2009} and
\verb!Vampire!~\cite{Riazanov1999}), high-order logic (e.g., \verb!Leo-II!
\cite{Benzmuller2008} and \verb!Satallax!~\cite{Brown2012}) or intuitionistic
logic (e.g., \verb!ileanCoP!~\cite{Otten2008}, \verb!JProver!
\cite{Schmitt2001}, and \verb!Gandalf!~\cite{Tammet1997}), among others.
On the other hand, we have the proof checkers, interactive theorem provers
(henceforth ITP) or proof assistants (e.g., \verb!Agda!~\cite{agdateam},
\verb!Coq!~\cite{coqteam}, \verb!Isabelle!~\cite{paulson1994isabelle}, and
\verb!HOL4!~\cite{norrish2007hol}).
The ITP tools provide us the logic framework to check and validate the
reply of the ATPs since they allow us to define the formal language for
the problems like operators, logic variables, axioms, and theorems.

A proof reconstruction tool provides such an integration in one direction,
is a translator of the reply delivered by the prover into the formalism of
the proof assistant. Because the formalism of the source (the evidence
generated by the ATP) is not necessarily the same logic in the target, the
reconstruction turns out in a ``reverse engineering'' task.
Then, reconstructing a proof involves an in-depth understanding of the
algorithms in the ATP and the specification logic in the ITP.
What we need from the ATP tools is a proof-object in a consistent format
to work with, that is, a full script describing step-by-step with
exhaustive details and without ambiguities, the table of the derivation
to get the actual proof. For problems in classical propositional logic
(henceforth CPL), from a list of at least forty\footnote{ATPs available
from the web service \texttt{SystemOnTPTP} of the TPTP World.} ATPs, just
a few provers were able to deliver proofs (e.g., \verb!CVC4!~
\cite{Barrett2011}, \verb!SPASS!, and
\verb!Waldmeister!~\cite{hillenbrand1997})
and a little bit less replied with a proof in a
file format like \verb!TSTP!~\cite{sutcliffe2004tstp} (e.g., \verb!E!, \verb!Metis!,
\verb!Vampire!, and \verb!Z3!), LFSC~\cite{Stump2008}, or the
SMT-LIB~\cite{Bohme2011} format.

Many approaches have been proposed and some tools have been implemented
for proof reconstruction in the last decades. These programs are relevant
not only because it helps to spread their usage but they also increase
the confidence of their users about their algorithms and their
correctness (see, for example, bugs in ATPs~\cite{Keller2013},
\cite{Bohme2011}, \cite{Fleury2014} and \cite{Kanso2012}).
% We mention some tools in the following.

In this paper, we describe the integration of \verb!Metis! prover with the
proof assistant \verb!Agda! by parsing \verb!TSTP! derivations to generate
\verb!Agda! proof-terms. We structure the paper as follows.
In section \ref{sec:metis-language-and-proofs} we briefly introduce
the \verb!Metis! prover. In section \ref{sec:proof-reconstruction}; we
present our approach to reconstruct proofs deliver by \verb!Metis! in
\verb!Agda!.
%In section \ref{sec:example}, we present a complete example of a proof reconstructed with our tool for a CPL problem.
We list related work in proof reconstruction in section \ref{sec:related-work}
In section~\ref{sec:conclusions}, we discuss some limitations, advantages,
and disadvantages with other similar proof reconstruction tools and
conclusions for ending up with the future work.

>>>>

% -*- root: main.tex -*-
\documentclass[../main.tex]{subfiles}
\begin{document}

@@

Proof reconstruction is a hard labor since it depends on the integration
of two complex systems. On the one hand, we have the automatic theorem provers
(henceforth \ATP) and their specification logic. These tools are usually
classified in at least one of the following categories. A \SAT solver
(e.g., \name{zChaff}~\cite{Moskewicz2001} and \name{MiniSat}~\cite{Een2004})
to prove unsatisfiability of \CNF formulas, a \abbre{QBF} solver
(e.g., \name{GhostQ}~\cite{Klieber2014} and \name{DepQBF}~\cite{Lonsing2017})
to prove satisfiability and invalidity of quantified Boolean formulas, a \SMT
solver (e.g \name{CVC4}~\cite{Barrett2011}, \name{veriT}~\cite{bouton2009},
and \name{Z3}~\cite{DeMoura2008}) to prove unsatisfiability of formulas
from first-order logic with theories, and a prover for validity of
formulas from first-order logic with equality
(e.g., \name{E}~\cite{Schulz:AICOM-2002}, \name{leanCoP}~\cite{Otten2008},
\Metis~\cite{hurd2003first}, \name{SPASS} ~\cite{Weidenbach2009} and
\name{Vampire}~\cite{Riazanov1999}), high-order logic (e.g., \name{Leo-II}
\cite{Benzmuller2008} and \name{Satallax}~\cite{Brown2012}) or intuitionistic
logic (e.g., \name{ileanCoP}~\cite{Otten2008}, \name{JProver}
\cite{Schmitt2001}, and \name{Gandalf}~\cite{Tammet1997}), among others.
On the other hand, we have the proof checkers, interactive theorem provers
(henceforth \ITP) or proof assistants (e.g., \Agda~\cite{agdateam},
\name{Coq}~\cite{coqteam}, \name{Isabelle}~\cite{paulson1994isabelle}, and
\name{HOL4}~\cite{norrish2007hol}).
The \ITP tools provide us the logic framework to check and validate the
reply of the \ATPs since they allow us to define the formal language for
the problems like operators, logic variables, axioms, and theorems.

A proof reconstruction tool provides such an integration in one direction,
is a translator of the reply delivered by the prover into the formalism of
the proof assistant. Because the formalism of the source (the evidence
generated by the \ATP) is not necessarily the same logic in the target, the
reconstruction turns out in a ``reverse engineering'' task.
Then, reconstructing a proof involves an in-depth understanding of the
algorithms in the \ATP and the specification logic in the \ITP.
What we need from the \ATP tools is a proof-object in a consistent format
to work with, that is, a full script describing step-by-step with
exhaustive details and without ambiguities, the table of the derivation
to get the actual proof. For problems in classical propositional logic
(henceforth \CPL), from a list of at least forty\footnote{\ATPs available
from the web service \name{SystemOnTPTP} of the TPTP World.} \ATPs, just
a few provers were able to deliver proofs (e.g., \name{CVC4}~
\cite{Barrett2011}, \name{SPASS}, and \name{Waldmeister}~\cite{hillenbrand1997})
and a little bit less replied with a proof in a file format like
\TSTP~\cite{sutcliffe2004tstp} (e.g., \name{E}, \Metis,
\name{Vampire}, and \name{Z3}), \len{LFSC}~\cite{Stump2008}, or the
\len{SMT-LIB}~\cite{Bohme2011} format.

Many approaches have been proposed and some tools have been implemented
for proof reconstruction in the last decades. These programs are relevant
not only because it helps to spread their usage but they also increase
the confidence of their users about their algorithms and their
correctness (see, for example, bugs in \ATPs~\cite{Keller2013},
\cite{Bohme2011}, \cite{Fleury2014} and \cite{Kanso2012}).
% We mention some tools in the following.

In this paper, we describe the integration of \Metis prover with the
proof assistant \Agda by parsing \TSTP derivations to generate
\Agda proof-terms. We structure the paper as follows.
In section \ref{sec:metis-language-and-proofs} we briefly introduce
the \Metis prover. In section \ref{sec:proof-reconstruction}; we
present our approach to reconstruct proofs deliver by \Metis in
\Agda.
We list related work in proof reconstruction in section \ref{sec:related-work}
In section~\ref{sec:conclusions}, we discuss some limitations, advantages,
and disadvantages with other similar proof reconstruction tools and
conclusions for ending up with the future work.

### 2017-09-05 paper/sections/metis-language-and-proofs.tex
[ paper ] new commands for programs and langs.

\verb!Metis! is an automatic theorem prover written in Standard ML for
first-order logic with equality developed by John Hurd~\cite{hurd2003first}.
This prover is suitable for proof reconstruction task since it provides
well-documented proof-objects with enough information to justify his deduction
steps and it also has been used by other proof-reconstruction tools.

\verb!Isabelle/HOL! integrated \verb!Metis! as a macro-step reconstruction
tool for explaining proof steps (usually CNF goals) replied from other ATPs like
\verb!CVC4!, \verb!Vampire!, or \verb!Z3!.
In \cite{Farber2015}, \verb!Metis! was used to provide a tactic for
\verb!HOL Light!; this tactic challenges others like \verb!MESON! tactic or the
\verb!leanCoP! tactic that reconstructs proofs delivered by
\verb!leanCoP!~\cite{Farber2016}.

In this paper, we ported a subset of \verb!Metis!' inference rules to \verb!Agda!,
the propositional fragmented, to allow us to justify step-by-step the proofs
delivered in \verb!TSTP! format generated by \verb!Metis!.

@@

\textit{Input.}~The \verb!TPTP! language  --which includes the first-order
form (denoted by \verb!fof!) and clause normal form (denoted by \verb!cnf!)
formats~\cite{sutcliffe2009} -- is de
facto input standard language to encode problems for many ATPs.
The \verb!TPTP! syntax describes a well-defined grammar to handle annotated
formulas with the following form

@@

The \verb!language! can be \verb!fof! or \verb!cnf!. The \verb!name!
serves to identify the formula within the problem. Each formula assumes one
\verb!role!, this could be an \verb!axiom!, \verb!conjecture!,
\verb!definition!, or a \verb!hypothesis!. The formulas include Boolean
connectives (\verb!&!, \verb!|!, \verb!=>!, \verb!<=>!, $\tt\sim$) and logic
constants \verb!$true! and \verb!$false!.

For instance, we express the problem
$p\, \vdash \neg (p \wedge \neg p) \vee (q \vee \neg q)$
in \verb!TPTP! syntax as follows.

@@

\textit{Output.}~\verb!TSTP! language is de facto output standard language
for derivations of ATPs~\cite{sutcliffe2004tstp}.
A \verb!TSTP! derivation is a directed acyclic graph, a proof tree, where each leaf
is a formula from the \verb!TPTP! input. A node is a formula inferred from the parent
formulas. The root is the final derived formula. Such a derivation is a list of
annotated formulas with the form

@@

The \verb!source! field is an inference record with the following pattern
\begin{verbatim}
inference(rule, useful info, parents).
\end{verbatim}

The \verb!rule! stands for the inference name; the other fields are
supporting arguments or useful information to apply the reasoning step, and
the list of parents nodes.
For instance, in Fig.~\ref{fig:metis-proof-tstp}, \verb!strip! is the name of one
of the inference with no arguments and only one parent node, the \verb!goal!.

\begin{figure}
\begin{verbatim}
fof(a, axiom, p) .
fof(goal, conjecture, p) .
fof(subgoal_0, plain, p, inference (strip, [], [goal])) .
fof(negate_0_0, plain, ~ p, inference (negate, [], [subgoal_0])) .
fof(normalize_0_0, plain, ∼ p,
  inference (canonicalize, [], [negate_0_0])) .
fof(normalize_0_1, plain, p,
  inference (canonicalize, [], [a])) .
fof(normalize_0_2, plain, $false,
  inference (simplify, [], [normalize_0_0, normalize_0_1]))
cnf(refute_0_0, plain, $false,
  inference (canonicalize, [], [normalize_0_2])) .
\end{verbatim}
\caption{A simple \texttt{Metis}' \texttt{TSTP} derivation for the problem $p\vdash p$.}
\label{fig:metis-proof-tstp}
\end{figure}

@@

A proof-object delivered in a \verb!Metis!' proof encodes a natural
deduction proof. The deduction system uses six inference rules
\cite{hurd2003first} (see Fig.~\ref{fig:metis-inferences}) to attempt to
prove conjectures by refutation (see. e.g., the last conclusion
in the TSTP derivation Fig.~\ref{fig:metis-example} is \emph{falsium}).

@@

These proofs are directed acyclic graphs, trees of refutations. Each node stands
for an application of an inference rule and the
leaves in the tree represent formulas in the given problem. Each node is
labeled with a name of the inference rule
(e.g., \verb!canonicalize! in Fig.~\ref{fig:metis-example}). Each
edge links a premise with one conclusion. The proof graphs have at their
root the conclusion $\bot$ since \verb!Metis! delivers proof by refutation.

\begin{figure}
\centering
\begin{bprooftree}\tt
  \AxiomC{}
  \RightLabel{\scriptsize negate}
  \UnaryInfC{$\neg p$}
  \RightLabel{\scriptsize strip}
  \UnaryInfC{$\neg p$}
  \AxiomC{}
  \RightLabel{\scriptsize axiom}
  \UnaryInfC{$p$}
  \RightLabel{\scriptsize canonicalize}
  \UnaryInfC{$p$}
  \RightLabel{\scriptsize simplify}
  \BinaryInfC{$\bot$}
  \RightLabel{\scriptsize canonicalize}
  \UnaryInfC{$\bot$}
\end{bprooftree}
% \caption{The \verb!Metis!' refutation tree for $p \vdash p$ from the
% derivation in Fig.~\ref{fig:metis-proof-tstp}}
\label{fig:metis-example}
\end{figure}

@@

Despite the six rules in the \verb!Metis!' logic
kernel (see~Fig. \ref{fig:metis-inferences}), we observed in \verb!TSTP!
derivations for CPL problems other inference rules.
They are \verb!canonicalize!, \verb!conjunct!, \verb!negate!, \verb!simplify!,
\verb!strip! and \verb!resolve!. We briefly review each of these rules
following no order except maybe by their frequency in the \verb!TSTP! derivations.

@@

\textit{Splitting}.
To prove a goal, \verb!Metis! splits the goal into disjoint cases. This
process produces a list of new subgoals, the conjunction of these subgoals
implies the goal as we show later in subsection~\ref{ssec:emulating-inferences}.
Then, a proof of the goal becomes in smaller proofs, one
refutation for each subgoal. These subgoals are introduced in the \verb!TSTP!
derivation with the \verb!strip! inference rule.

@@

\textit{Clausification.} The \verb!clausify! rule transforms a
propositional formula in clausal normal form or CNF,
a conjunction of clauses. Where a clause is the disjunction of zero or
more literals and a literal is an atom (positive literal) or a negation of an
atom (negative literal). This conversion is not unique, and \verb!Metis! performs
this transformation using its criteria.

@@

\textit{Normalization.}
\verb!Metis! uses the \verb!canonicalize! rule to normalize a
propositional formula that comes from an axiom or a definition.
It often converts each expression in one of its normal form,
the conjunctive normal form, the negative normal form or the disjunctive normal form.
Posterior, this rule simplifies the formula with some definitions, applying
recursively on the structure's formula conjunctions and disjunctions simplifications
to remove tautologies
(see Fig.~\ref{fig:conjunctive-disjunctive-simplification} with a list
of these theorems. We had assumed the commutative property for conjunction and
disjunction connectives).

@@

\[%\scalebox{0.9}{
  \begin{bprooftree}
    \AxiomC{$\varphi \vee \bot$}
    \UnaryInfC{$\varphi$}
  \end{bprooftree}
  \qquad
  \begin{bprooftree}
    \AxiomC{$\varphi \vee \top$}
    \UnaryInfC{$\top$}
  \end{bprooftree}
  \qquad
  \begin{bprooftree}
    \AxiomC{$\varphi \vee \neg \varphi$}
    \UnaryInfC{$\top$}
  \end{bprooftree}
  \qquad
  \begin{bprooftree}
    \AxiomC{$\varphi \vee \varphi$}
    \UnaryInfC{$\varphi$}
  \end{bprooftree}
%}
\]
% \caption{Some rules of \verb!canonicalize! inference.}
\label{fig:conjunctive-disjunctive-simplification}
\end{figure}

@@

\textit{Resolution.} The \verb!resolve! rule is the resolution
theorem showed in Fig.~\ref{fig:metis-inferences}. Its applications
needs the positive literal for resolution and two
derivations to deduce the \emph{resolvent}.
The positive literal $p$ must occur in the formula from the first derivation,
and the negative literal must occur in the formula from the second derivation
(this follows the same pattern in \verb!Metis!' logic core in
Fig.~\ref{fig:metis-inferences} for the resolve inference).

@@

\textit{Splitting a conjunct.}
The \verb!conjunct! rule extracts from a conjunction one of its conjuncts; it is
a generalization of the projection rules for the conjunction connective as
we can see in the following \verb!TSTP! derivation.

@@

\textit{Negate.}
Each subgoal proof is a refutation, thereby each proof assumes
the negation of its subgoal. The \verb!negate! rule
introduces the negation of a subgoal that results
after applies the \verb!strip! inference to the goal.

@@

\textit{Simplification.} %abbreviate formulas
The \verb!simplify! rule is a $n$-ary inference that performs simplification
of definitions. This rule transverses a list of formulas by
applying different theorems (e.g., \emph{modus pones}, \emph{modus tollens},
or \emph{disjunctive syllogism}) including the list of theorems used by
\verb!canonicalize! in Fig.~\ref{fig:conjunctive-disjunctive-simplification}
and the resolution theorem of \verb!resolve! rule.
Moreover, many things happen inside this \verb!Metis!' procedure, and
it is somewhat complicated to grasp it completely.

>>>>

\Metis is an automatic theorem prover written in \len{Standard ML} for
first-order logic with equality developed by John Hurd~\cite{hurd2003first}.
This prover is suitable for proof reconstruction since it provides
well-documented proof-objects to justify its deduction steps from the basis
of only six inference rules. Exists tools that perform proof
reconstruction with \Metis, some of these systems are the following.

\name{Isabelle/HOL} incorporated \Metis as a macro-step reconstruction
tool for explaining proof steps (usually \CNF goals) replied from other \ATPs like
\name{CVC4}, \name{Vampire}, or \name{Z3}
(more details in Ref.~\cite{paulson2007source}).
In \cite{Farber2015}, \Metis was used to provide a tactic for
\name{HOL Light}; this tactic challenges others like \name{MESON} tactic or the
\name{leanCoP} tactic that reconstructs proofs delivered by
\name{leanCoP}~\cite{Farber2016}.

In this paper, we ported a subset of \Metis' inference rules to \Agda,
the propositional fragmented, to allow us to justify step-by-step the proofs
delivered in \TSTP format generated by \Metis.

@@

\textit{Input.}~The \TPTP language --which includes the first-order
form (denoted by \name{fof}) and clause normal form (denoted by \name{cnf})
formats~\cite{sutcliffe2009} -- is de
facto input standard language to encode problems for many \ATPs.
The \TPTP syntax describes a well-defined grammar to handle annotated
formulas with the following form

@@

The \name{language} can be \name{fof} or \name{cnf}. The \name{name}
serves to identify the formula within the problem. Each formula assumes one
\name{role}, this could be an \name{axiom}, \name{conjecture},
\name{definition}, or a \name{hypothesis}. The formulas include Boolean
connectives (\verb!&!, \verb!|!, \verb!=>!, \verb!{<=>!, $\tt\sim$) and logic
constants \verb!$true! and \verb!$false!.

For instance, we express the problem
$p\, \vdash \neg (p \wedge \neg p) \vee (q \vee \neg q)$
in \TPTP syntax as follows.

@@

\textit{Output.}~\TSTP language is de facto output standard language
for derivations of ATPs~\cite{sutcliffe2004tstp}.
A \TSTP derivation is a directed acyclic graph, a proof tree, where each leaf
is a formula from the \TPTP input. A node is a formula inferred from the parent
formulas. The root is the final derived formula. Such a derivation is a list of
annotated formulas with the form

@@

The \name{source} field is an inference record with the following pattern
\begin{verbatim}
inference(rule, useful info, parents).
\end{verbatim}

The \name{rule} stands for the inference name; the other fields are
supporting arguments or useful information to apply the reasoning step, and
the list of parents nodes.
For instance, in Fig.~\ref{fig:metis-proof-tstp}, \strip is the name of one
of the inference with no arguments and only one parent node, the \name{goal}.

\begin{figure}
\begin{verbatim}
fof(a, axiom, p) .
fof(goal, conjecture, p) .
fof(subgoal_0, plain, p, inference (strip, [], [goal])) .
fof(negate_0_0, plain, ~ p, inference (negate, [], [subgoal_0])) .
fof(normalize_0_0, plain, ∼ p,
  inference (canonicalize, [], [negate_0_0])) .
fof(normalize_0_1, plain, p,
  inference (canonicalize, [], [a])) .
fof(normalize_0_2, plain, $false,
  inference (simplify, [], [normalize_0_0, normalize_0_1]))
cnf(refute_0_0, plain, $false,
  inference (canonicalize, [], [normalize_0_2])) .
\end{verbatim}
\caption{The \texttt{Metis}' \texttt{TSTP} derivation for the problem $p\vdash p$.}
\label{fig:metis-proof-tstp}
\end{figure}

@@

A proof-object delivered in a \Metis' proof encodes a natural
deduction proof. The deduction system uses six inference rules
\cite{hurd2003first} (see Fig.~\ref{fig:metis-inferences}) to attempt to
prove conjectures by refutation (see. e.g., the last conclusion
in the \TSTP derivation Fig.~\ref{fig:metis-example} is \emph{falsium}).

@@

These proofs are directed acyclic graphs, trees of refutations. Each node stands
for an application of an inference rule and the
leaves in the tree represent formulas in the given problem. Each node is
labeled with a name of the inference rule
(e.g., \canonicalize in Fig.~\ref{fig:metis-example}). Each
edge links a premise with one conclusion. The proof graphs have at their
root the conclusion $\bot$ since \Metis delivers proof by refutation.

\begin{figure}
\centering
\begin{bprooftree}\tt
  \AxiomC{}
  \RightLabel{\scriptsize negate}
  \UnaryInfC{$\neg p$}
  \RightLabel{\scriptsize strip}
  \UnaryInfC{$\neg p$}
  \AxiomC{}
  \RightLabel{\scriptsize axiom}
  \UnaryInfC{$p$}
  \RightLabel{\scriptsize canonicalize}
  \UnaryInfC{$p$}
  \RightLabel{\scriptsize simplify}
  \BinaryInfC{$\bot$}
  \RightLabel{\scriptsize canonicalize}
  \UnaryInfC{$\bot$}
\end{bprooftree}
% \caption{The \Metis' refutation tree for $p \vdash p$ from the
% derivation in Fig.~\ref{fig:metis-proof-tstp}}
\label{fig:metis-example}
\end{figure}

@@

Despite the six rules in the \Metis' logic
kernel (see~Fig. \ref{fig:metis-inferences}), we observed in \TSTP
derivations for\CPL problems other inference rules.
They are \canonicalize, \conjunct, \negate, \simplify,
\strip and \resolve. We briefly review each of these rules
following no order except maybe by their frequency in the \TSTP derivations.

@@

\textit{Splitting}.
To prove a goal, \Metis splits the goal into disjoint cases. This
process produces a list of new subgoals, the conjunction of these subgoals
implies the goal as we show later in subsection~\ref{ssec:emulating-inferences}.
Then, a proof of the goal becomes in smaller proofs, one
refutation for each subgoal. These subgoals are introduced in the \TSTP
derivation with the \strip inference rule.

@@

\textit{Clausification.} The \clausify rule transforms a
propositional formula in clausal normal form or \CNF,
a conjunction of clauses. Where a clause is the disjunction of zero or
more literals and a literal is an atom (positive literal) or a negation of an
atom (negative literal). This conversion is not unique, and \Metis performs
this transformation using its criteria.

@@

\textit{Normalization.}
\Metis uses the \canonicalize rule to normalize a
propositional formula that comes from an axiom or a definition.
It often converts each expression in one of its normal form,
the conjunctive normal form, the negative normal form or the disjunctive normal form.
Posterior, this rule simplifies the formula with some definitions, applying
recursively on the structure's formula conjunctions and disjunctions simplifications
to remove tautologies
(see Fig.~\ref{fig:conjunctive-disjunctive-simplification} with a list
of these theorems. We had assumed the commutative property for conjunction and
disjunction connectives).

@@

\[%\scalebox{0.9}{
  \begin{bprooftree}
    \AxiomC{$\varphi \vee \bot$}
    \UnaryInfC{$\varphi$}
  \end{bprooftree}
  \qquad
  \begin{bprooftree}
    \AxiomC{$\varphi \vee \top$}
    \UnaryInfC{$\top$}
  \end{bprooftree}
  \qquad
  \begin{bprooftree}
    \AxiomC{$\varphi \vee \neg \varphi$}
    \UnaryInfC{$\top$}
  \end{bprooftree}
  \qquad
  \begin{bprooftree}
    \AxiomC{$\varphi \vee \varphi$}
    \UnaryInfC{$\varphi$}
  \end{bprooftree}
%}
\]
% \caption{Some rules of \canonicalize inference.}
\label{fig:conjunctive-disjunctive-simplification}
\end{figure}

@@

\textit{Resolution.} The \resolve rule is the resolution
theorem showed in Fig.~\ref{fig:metis-inferences}. Its applications
needs the positive literal for resolution and two
derivations to deduce the \emph{resolvent}.
The positive literal $p$ must occur in the formula from the first derivation,
and the negative literal must occur in the formula from the second derivation
(this follows the same pattern in \Metis' logic core in
Fig.~\ref{fig:metis-inferences} for the resolve inference).

@@

\textit{Splitting a conjunct.}
The \conjunct rule extracts from a conjunction one of its conjuncts; it is
a generalization of the projection rules for the conjunction connective as
we can see in the following \TSTP derivation.

@@

\textit{Negate.}
Each subgoal proof is a refutation, thereby each proof assumes
the negation of its subgoal. The \negate rule
introduces the negation of a subgoal that results
after applies the \strip inference to the goal.

@@

\textit{Simplification.} %abbreviate formulas
The \simplify rule is a $n$-ary inference that performs simplification
of definitions. This rule transverses a list of formulas by
applying different theorems (e.g., \emph{modus pones}, \emph{modus tollens},
or \emph{disjunctive syllogism}) including the list of theorems used by
\canonicalize in Fig.~\ref{fig:conjunctive-disjunctive-simplification}
and the resolution theorem of \resolve rule.
Moreover, many things happen inside this \Metis' procedure, and
it is somewhat complicated to grasp it completely.
